Python as a machine learning language: experiences from the scikit-learn project

### 2013-06-12 paper.tex
typo
trim/rewrite the old weaknesses section to fit in an implementation section

Finally, let us also note that, the \texttt{fit} method always returns the estimator
it was called on, hence making method chaining possible.  Using this technique,
we can thus rewrite the \texttt{StandardScaler} example above in a single line:
\begin{verbatim}
    X_train = StandardScaler().fit(X_train).transform(X_train)
\end{verbatim}
In addition, \texttt{fit(X\_train).transform(X\_train)} can further
be rewritten as \texttt{fit\_transform(X\_train)}.  Combining the \texttt{fit}
and \texttt{transform} methods into a single method \texttt{fit\_transform}
allows to share computations, which often makes a huge difference in practice.
In the same spirit, \sklearn sometimes provides the
\texttt{fit\_predict} method for combining the \texttt{fit} and \texttt{predict}
methods. This is especially useful for retrieving cluster predictions with
respect to the training data \texttt{X\_train} in clustering algorithms.

@@

% This needs an introductory paragraph, but the best I could come up with was:
As can be expected, the NumPy/SciPy environment is not completely flawless.

% The following hedge needs to be in here; we don't want to diss our peers
It should be noted that this section is meant
as a summary of challenges faced by scikit-learn development
and an attempt at constructive criticism.
There is considerable overlap between the communities
that develop or contribute to NumPy, SciPy, scikit-learn, and Cython,
so the issues raised here may eventually be resolved
by the scikit-learn developers themselves.

\subsection{Sparse matrix support}

SciPy's handling of sparse matrices is one source of confusion
for scikit-learn developers and users alike.
The \texttt{scipy.sparse} package offers support
for this kind of data structure,
which is crucial to achieving performance and scalability
in the face of high-dimensional inputs
(including, but not limited to, NLP applications).
However, the interface exposed by its classes
deviates from the NumPy array interface, sometimes subtly so,
sometimes entirely.  % XXX is this grammatically correct?
For example, while NumPy arrays have a variable number of dimensions,
so that vectors can be considered 1-d arrays
(without a distinction between row and column vectors),
matrices are 2-d, etc., \texttt{scipy.sparse} matrices are always 2-dimensional,
meaning that special care has to be taken when handling sparse vectors.

The development team has built up sufficient familiarity
with the sparse/array distinction to factor out common patterns
for supporting both sparse and dense input data structures in many estimators.
In other cases, though, sparse matrix support has to implemented
on a per-estimator basis.

@@

At this point in time, support for multi-core programming in Python,
at least in the CPython reference implementation targeted by scikit-learn,
is downright weak.
The interpreter supports multithreading, but due to the ``Global Interpreter Lock'' (GIL),
no two threads in a process may at the same time execute Python code.
It follows that multithreading is not a viable solution
even two embarrassingly parallel problems,
such as fitting and evaluation multiple models in a grid search procedure.
The workaround is to use Python's multiprocessing module,
which emulates threads using multiple operating system-level processes,
but this in turn leads to high memory use and overhead
because of the copying of training and test sets
into the various processes' address spaces.

While Cython makes writing performance-critical code possible
in a language that is very much like Python,
doing so properly still requires some knowledge of C.
One source of trouble has been the various fixed-size integer types
that are used by the Python and NumPy C runtimes,
whose sizes are practically all platform-dependent.

Another problem concerns the mismatch between on the one hand Python functions,
which are first-class runtime objects endowed with methods and metadata
that identify them as ``callable'',
that dynamically check their number and types of arguments,
and that can return any Python type of object,
and compiled C or Cython functions on the other hand,
which exhibit static typing.
Where it may sometimes be convenient for estimators
to take functions as arguments
(e.g., in the case of kernel machine learning,
where one might like to use a custom kernel),
but such usage is not at present possible and may never be.
Passing C or Cython-compiled functions would require an API for the kernel method
to be implemented at the C or Cython level,
only to be used from compiled code.
Passing arbitrary Python functions would be possible,
but performance can be expected to degrade significantly
because of the runtime checks needed for each invocation of the kernel function.
(The current solution in scikit-learn is to let the user pass not a kernel function,
but a precomputed Gram matrix for an entire set of objects.
The linear, polynomial and RBF kernels are treated as special cases in
the \textsf{LIBSVM}-based SVM learner, where they are implemented in C{}\verb!++!.)

>>>>

Finally, let us also note that the \texttt{fit} method always returns the estimator
it was called on, hence making method chaining possible.  Using this technique,
we can thus rewrite the \texttt{StandardScaler} example above in a single line:
\begin{verbatim}
    X_train = StandardScaler().fit(X_train).transform(X_train)
\end{verbatim}
In addition, \texttt{fit(X\_train).transform(X\_train)} can further
be rewritten as \texttt{fit\_transform(X\_train)}.  Combining the \texttt{fit}
and \texttt{transform} methods into a single method \texttt{fit\_transform}
allows to share computations, which often makes a huge difference in practice.
In the same spirit, \sklearn sometimes provides the
\texttt{fit\_predict} method for combining the \texttt{fit} and \texttt{predict}
methods. This is especially useful for retrieving cluster predictions with
respect to the training data \texttt{X\_train} in clustering algorithms.

@@

@@

At this point in time, support for multi-core programming in Python,
at least in the CPython reference implementation targeted by scikit-learn,
is suboptimal.
The interpreter supports multithreading,
but due to the ``Global Interpreter Lock'' (GIL),
no two threads in a process may at the same time execute Python code.
It follows that multithreading is not a viable solution
even for embarrassingly parallel problems,
such as fitting and evaluation multiple models in a grid search procedure.
The workaround is to use Python's multiprocessing module,
which emulates threads using multiple operating system-level processes,
but this in turn leads to high memory use and overhead
because of the copying of training and test sets
into the various processes' address spaces.

\subsection{Cython for inner loops}

A problem arising from the use of Cython
concerns the mismatch between on the one hand Python functions,
which are first-class runtime objects endowed with methods and metadata
that identify them as ``callable'',
that dynamically check their number and types of arguments,
and that can return any Python type of object;
and compiled C or Cython functions on the other hand,
which exhibit static typing.
Where it may sometimes be convenient for estimators
to take functions as arguments
(e.g., in the case of kernel machine learning,
where one might like to use a custom kernel),
but such usage is not at present possible and may never be.
Passing C or Cython-compiled functions
would require an API for the kernel method
to be implemented at the C or Cython level,
only to be used from compiled code.
Passing arbitrary Python functions would be possible,
but performance can be expected to degrade significantly
because of the runtime checks needed for each invocation of the kernel function.
(The current solution in scikit-learn is to let the user pass
not a kernel function,
but a precomputed Gram matrix for an entire set of objects.
The linear, polynomial and RBF kernels are treated as special cases in
the \textsf{LIBSVM}-based SVM learner,
where they are implemented in C{}\verb!++!.)
